---
title: "Statistical Inference Course Project Part 2"
author: "JB"
date: "2023-10-04"
output:
  pdf_document: default
  html_document: default
---
# Overview
This report  shows basic Inferential Data analysis on the ToothGrowth data set.

```{r}
# Loading necessary packages
library(datasets)
library(ggplot2)
library(knitr)
library(dplyr)
library(tidyr)
```

## Part 2: Basic Inferential Data Analysis.
```{r}
data("ToothGrowth")
dt<- ToothGrowth
unique(dt$dose)
```
There are three unique values for dose which I will convert to factors.
```{r}
dt$dose<- factor(dt$dose)
str(dt)
```

 1. Plotting box plot to visualise the difference between two supplement types and dosage.
```{r}
ggplot(dt, aes(x=dose, y=len, fill = supp))+
          geom_boxplot()+
          ggtitle("Tooth growth length based on supplement type and dose")+
          theme_classic()
```

From the plot we see that for the first two doses there seem to be an observable difference in tooth growth between the two supplemtents. We also see that with increased dose the tooth growth length increases. Last conclusiion we can make is that for 0.5 and 1 dose OJ supplememnt seems to be performing better whereas for 2 dose there isnt much difference between OJ and VC.

 2. Calculating mean difference between supplements for all three doses.
```{r}
dt %>% 
          group_by(supp, dose)%>%
          summarise(mean = mean(len), .groups = "drop")%>%
          spread(supp, mean) %>%
          mutate(diff = abs(VC-OJ))
```
Only for dose 2 the difference between two supplememnts is really smallmeaning its harder to comapre their effectiveness.

 3. T test hypothesis for all doses.

Null hypothesis is that there is no significant difference between OJ and VC. Althernative hypothesis says there is a difference between the two drugs. We set alpha for standard 0.05.
```{r}
# Filtering data for testing
dose_half<- filter(dt, dose==0.5)
dose_one<- filter(dt, dose==1)
dose_two<- filter(dt, dose==2)
# t-test for 0.5 dose
t.test(len~supp, dose_half)
# t-test for 1 dose
t.test(len~supp, dose_one)
# t-test for 2 dose
t.test(len~supp, dose_two)

# Table summarising the tests results
table_data <- data.frame(
          Dose = c(0.5, 1, 2),
          p_value = c(0.006359, 0.001038, 0.9639),
          Conf.Int = c("1.719057 8.780943", "2.802148 9.057852", "-3.79807 3.63807"),
          Decision = c("Reject Null", "Reject Null", "Do not Reject Null")
)

print(table_data)
```
**Conclusion** As expected the p-value for dose two is significantly bigger than for dose 1 and 0.5 meaning we cannot reject the hypothesis that supplements OJ and VC are different. We can't strictly say that those two supplements have a different effect on tooth growth. The conclusion is made under assumption that the data isn't paired and the variance are different.
