---
title: "Statistical Inference Course Project Part 1"
author: "JB"
date: "2023-10-04"
output:
  pdf_document: default
  html_document: default
---
# Overview
This report investigates the exponential distribution inR in comparison to Central Limit Theorem  shows basic Inferential Data analysis on the ToothGrowth data set in part 2.
```{r}
# Loading necessary packages
library(datasets)
library(ggplot2)
library(knitr)
library(dplyr)
library(tidyr)
```
## Part 1: Simulation exercise.
This exercise starts with generating 1000 simulations of 40-size samples of exponential distribution. The simulation is done with a rexp() function in R. The mean of each simulation is calculated using the apply() function.
```{r}
lambda<- 0.2
n<- 40
simData<- replicate(1000, rexp(n,lambda))

simMeans<- apply(simData, 2, mean)
```
 
 1. Show the sample mean and compare it to the theoretical mean of the distribution.
```{r}
sampleMean<- mean(simMeans)
sampleMean
theoreticalMean<- 1/lambda
theoreticalMean
```

 2. Plot Histogram showing the distribution of 1000 mean values calculated for each simulation.
```{r}
hist(simMeans, main = "Figure1. Comparison of Sample Meand and Theoretical Mean", xlab="Mean")
abline(v= sampleMean, lw = 4, col= "red")
abline(v= theoreticalMean, lw = 4, col = "green")
```

The red line indicates the simulated data mean, whereas the green the theoretical one.

 3. Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.
```{r}
sampleSD<- sd(simMeans)
sampleSD
theoreticalSD<- (1/lambda)/sqrt(n) 
theoreticalSD
sampleVar<- sampleSD^2
sampleVar
theoreticalVAr<- theoreticalSD^2
theoreticalVAr
```
 
 4. Show that the distribution is approximately normal.
```{r}
xfit<- seq(min(simMeans), max(simMeans), length=100)
yfit<- dnorm(xfit, mean = 1/lambda, sd= (1/lambda)/sqrt(n))
hist(simMeans, breaks=n, prob = T, main = "Density of means", xlab = "Means", ylab = "Density" )
lines(xfit, yfit, pch = 22, lw=3, lty = 5, col="red")
```

The red curve indicates the normal distribution

Summary between the two distribution can be seen below
```{r, echo=FALSE}
table_data <- data.frame(
          Variable = c("Mean", "Standard Deviation", "Variance"),
          Simulated = c(sampleMean, sampleSD, sampleVar),
          Theoretical = c(theoreticalMean, theoreticalSD, theoreticalVAr)
)
print(table_data)
```

**Conclucion** As we can see from the table all of mean, standard deviation and variance are closely similar. Due to the CLT the distribution of averages of 40 exponential is closely similar to a normal distribution. 

